### 2013-01-11 manuscript.tex
adicionei as modificações

Most if not all existing lossless compression methods were developed under the following usage paradigm: \textit{produce} $\rightarrow$ \textit{compress} $\rightarrow$ \textit{store} $\rightarrow$ \textit{uncompress} $\rightarrow$ \textit{use}. The focus of the present work is to allow a slightly different usage:  \textit{produce} $\rightarrow$ \textit{compress} $\rightarrow$ \textit{perform mathematical manipulations} $\rightarrow$ \textit{decompress} (only for human reading).

@@

This paper will focus solely on numerical data which for the purpose of the applications is organized as   matrices. This is a most common data structure found in computational data analysis environments. The matrices compressed according to the methodologies proposed here should be able to undergo the same mathematical operations as the original uncompressed matrices. e.g. linear algebra manipulations. This way, the cost of compression is reduced to a single event of compression and no need of decompression except when displaying the results for human reading. The idea of operating with compressed arrays is relatively new\cite{yemliha2007compiler}, and has yet to find mainstream applications to the field of numerical computations. One application which employs a form of compression is the sparse matrix linear algebra algorithms\cite{dodson1991sparse}, in this case there is no alteration in the standard encoding of the data, but only the non-zero elements of the matrices are stored and operated upon.

@@

To maintain mathematical equivalence with the original data for any arithmetic operations, we need to  maintain the structure of the matrix, i.e., the ability to acess any element given its row $i$ and column $j$ and also the numeric nature of its elements. In order to achieve compression we decided to exploit inefficiencies in the conventional way matrices are allocated in memory. The exemples in this paper will be restricted to matrices with positive integer elements.

The compression method is as follows. Let $M_{r \times c}$ be a matrix, in which $r$ is the number of rows and $c$ the number of columns. Each element of this matrix, called $m_{ij}$, is a positive integer. In digital computers, all information is stored as binary code (base 2 numbers). However the conventional way to store arrays of integers is on a sequence of memory blocks of fixed size (power of 2 number of bits), one for each element. The maximum size of a block is equal to the word size of the processor, which for most current CPUs is 64 bits. Some special number such as complex number may be encoded as two blocks instead of one. The size of the chunk of memory allocated to each number will determine their maximum  size (for integers) or their precision (for floating-point numbes). So for matrix $M$, the total memory allocated, assuming chunks of 64 bits, is given by $\mathcal{B} = r \times c \times 64$.

@@

The SM method consists in determining the value of the greatest element of matrix $M$, which coincides with its supremum, $max M == \sup M$  and determine the minimum number of bits, $b(\sup M$), required to store it. We will use capital roman letters to denote uncompressed matrices and the corresponding lower case letter for the compressed version.

\begin{equation} \label{eq:07}
 b(\sup M) \approx \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 1  
	\end{cases}
\end{equation}

The allocation of memory still happens in the usual way, i.e. in fixed size 64-bit chunks, only that now, in the space required for a single 64 bit integer, we can store for example, an entire $8\times 8$ matrix of $0_{10}$ and $1_{10}$.

Let's look at a concrete example: suppose that the greatest value to be stored in a matrix $M$ is $\max M=1023$. Therefore, the number of bits required to represent it is $10$ $(1111111111)$. Let the first 8 elements of $M$ be:

\begin{equation} \label{eq:08}
  M = \begin{bmatrix}
  900 & 1023 & 721 & 256 & 1 & 10 & 700 & 20 & \hdots\\ 
  \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \ddots 
  \end{bmatrix}
\end{equation}

These elements of $M$, in binary, are shown in Table \ref{tab:01}. It is evident that the number of bits required to represent any other element must be $\leq 10$. From now on the minimum number of bits required to represent a base 10 integer will be refered to as its bit-length.

@@

to store matrix $M$ it first has to be converted to base $2$ ($M_2$). Then it will be unraveled by column (column major, e.g. in Fortran) or by row (row major, e.g. in C) and its elements will be written as  fixed size adjacent chunks of memory. The size of each chunk is determined by the type associated with the matrix (typically 64 bits, but always a power of 2).

According to the SM method, having determined that each element will require at most 10 bits, we can divide the memory block corresponding to a single 64 bit integer into 6 10-bit chunks which can each hold a single element of $M$. These 64-bit blocks will be called a bitstring. The remaining 4 bits will be used later. The number of bitstrings needed will be $\lfloor \frac{\dim(M) *b(\sup M)} {64} \rfloor +1$.

The final layout of the first 6 elements of $m$ in the first bitstring can be seen in \ref{eq:09}.

\begin{equation}\label{eq:09}
 bitstring_1 = 0000\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\underbrace{0100000000}_{256}\underbrace{1011010001}_{721}\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
\end{equation}

Here is a step-by-step description of the aplicationof the SM method to matrix $M$:
\begin{enumerate}
 \item Element $M_{1,1}=900 = m_{1,1} = 1110000100$ is stored in the first 10-bit chunk of the element strip $bitstring[1]$ , which corresponds to bits 0 to 9.
 \item Element $M_{1,2}=1023$ is allocated in the second chunk, from bit 10 to bit 19.
 \item Repeat for elements $M_{1,i}$ with $i=1,\ldots,6$ which are stored on the remaining chunks.
 \begin{equation*}\label{eq:10}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{b}}_{bitstring_2}|\underbrace{\overbrace{1100}^{a}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{bitstring_1}
\end{equation*}
 \item Element 7 does not fit on the remaining 4 bits of the first bitstring. So it will straddle two bitstrings, i.e., it is divided in two segments $a$ and $b$, $a$ is written on the first bitstring $b$ on the second.
 \begin{equation*}\label{eq:10}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{b}}_{bitstring_2}|\underbrace{\overbrace{1100}^{a}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{bitstring_1}
\end{equation*}
\end{enumerate}

@@

\begin{equation*}\label{eq:11}
  \underbrace{0000010100}_{20}\underbrace{\overbrace{101011}^a|\overbrace{1100}^{b}}_{700}\underbrace{0000001010}_{10}\ldots
\end{equation*}

@@

Lets use again the matrix described in equation \ref{eq:08}, where  the largest element is number 1023. Now instead of assigning one chunk of the bitstring to each element of $m$, we will assign two chunks: the first will store the number of bits required to store the element and the second  will store the actual element. The first chunk will have a fixed size, in this case, 4 bits. These 4 bits are the required space to store the bit-length of $\sup M $, in this case, 10.

@@

\begin{enumerate}
 \item The first element $M_{1,1}=900$ requires 10 bits to store, so we write $10$ in the first chunk and $900$ in the second.
 \begin{equation*} \label{eq:13}
 bitstring_1 = 00000000000000000000000000000000000000000000000000\underbrace{\underbrace{1110000100}_{\text{element} = 900}\underbrace{1010}_{\text{bit-length=10}}}_{M_{1,1}}
\end{equation*}
 \item Do the same for the next element, $M_{1,2}=1023$.
 \item Element $M_{1,2}=721$ is also added taking the bitstring to the state.
 \begin{equation*} \label{eq:14}
 bitstring_1 = 
 0000000000000000000000\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10}
\end{equation*}
\end{enumerate}

@@

\begin{enumerate}
 \item[4.] Element $M_{1,4} =256$ is added.  
 \item [5.] Elements $M_{1,5} =1$ and $M_{1,6} =10$ are added requiring a total of 13 bits instead of 20 with the SM method.
 \begin{align*} \label{eq:16}
 bitstring_1 &=
 \underbrace{0100}_{4}\underbrace{1}_{1}\underbrace{0001}_{1}\underbrace{100000000}_{256}\underbrace{1001}_{9}\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10} \\
 bitstring_2 &= 000000000000000000000000000000000000000000000000000000000000\underbrace{1010}_{10}
\end{align*}
 \item [6.] The remaining two elements are added $M_{1,7} =700$ and $M_{1,8} =20$.
 \begin{equation*} \label{eq:18}
 bitstring_2 = 00000000000000000000000000000000000\underbrace{10100}_{20}\underbrace{0101}_{5}\underbrace{1010111100}_{700}\underbrace{1010}_{10}\underbrace{1010}_{10}
\end{equation*}
\end{enumerate}

@@

 \begin{equation}\label{eq:eff1}
  \eta=\frac{bits\, alocated-bits\, used}{bits\,alocated}
 \end{equation}

 Where \textit{bits alocated} above mean total bits required for standard storage of the matrix, without compression, while bits used mean total bits requires to store the matrix after compression. From nowon the efficiencies are denoted by $\eta_1$ for the SM method and by $\eta_2$ for the VLB method. 
 \subsubsection*{SM Method}
 Let $M_{r \times c}$ be the matrix we wish to compress. In comparison with a conventional allocation (64-bit integers) we can apply equation \ref{eq:eff1} to caluculate the efficiency of the SM method:

\begin{align}\label{eq:22}
 \eta_1 &= \frac{64 \times rc - b(max M) \times rc}{64 \times rc}\nonumber\\
 &= \frac{64  - b(max M) }{64}
\end{align}

As we see in \ref{eq:22}, $\eta_1$ does not depend on size of the matrix, only on the bit-length of $\max M$. If $b(\max M)=64$, $\eta_1$ is 0, i. e. no compression is possible. On the other extreme, if the matrix is composed exclusively of 0s and 1s, maximal compression is achievable, $\eta_1=1$.

@@

\begin{equation}\label{eq:24}
 \eta_2 = \frac{64 \times rc - \sum_{i=1}^{g} ( b_i + k ) \times f_i }{64 \times rc} 
\end{equation}

We can further simplify equation \ref{eq:24} to get at shorter expression for the compression ratio.

\begin{align} \label{eq:25}
 \eta_2 &= \frac{64 \times rc - \sum_{i=1}^{g} ( b_i \times f_i + k \times f_i )}{64 \times rc} \nonumber \\
  &= \frac{64 \times rc - \sum_{i=1}^{g}  b_i \times f_i  -\sum_{i=1}^{g}  k \times f_i }{64 \times rc}\nonumber \\
  &= \frac{64 \times rc - \sum_{i=1}^{g}  b_i \times f_i  - k \times\sum_{i=1}^{g}  f_i }{64 \times rc}\nonumber \\
 &\sum_{i=1}^{g} f_i = rc \nonumber \\
 \eta_2 &= \frac{64 \times rc - \sum_{i=1}^{g}  b_i \times f_i  - k \times rc }{64 \times rc}\nonumber \\
  &= 1 - \frac{\sum_{i=1}^{g}  b_i \times f_i }{64 \times rc} - \frac{k}{64}
\end{align}

@@

\paragraph{Random Matrix Generation.}
In both methods, compression eficiency depends on the distribution of the bit-lengths $b(m_{i,j})$. Thus, in this section, a method to generate a variety of random bit-length distributions is proposed.

For simplicity we will model the distribution of  as a mixture $X$ of two Beta distributions ,$ B_1 \sim Beta(\alpha_1,\beta_1)$ and $B_2 \sim Beta(\alpha_2,\beta_2)$, whose probability function is shown in equation (eq.\ref{eq:mixture}). Since the Beta distribution is defined only in the interval $[0,1] \subset \mathbb{R}$ , we applied a simple transformation ($\lfloor 64 \times x \rfloor + 1$) to the mixture in order to map it to the interval of $[1,64] \subset \mathbb{Z}$

\begin{equation}\label{eq:mixture}
 F(x) =  w \, Beta(\alpha_1,\beta_1) + (1-w)\,Beta(\alpha_2,\beta_2)
\end{equation}

The intention of using this mixture was to find a simple way to represent a large variety of bit length distributions. The first two central moments of this mixture are given in \ref{eq:53} and will be used later to summarize our numerical results.

\begin{align}\label{eq:53}
 E(X) &= w E(B_1) + (1-w) E(B_2) \nonumber \\
 &= w \frac{\alpha_1}{\alpha_1+\beta_1} + (1-w)\frac{\alpha_2}{\alpha_2+\beta_2}\nonumber \\
 Var(X) &= w Var(B_1) + (1-w) Var(B_2) + w (1-w) (E(B_1)^2 - E(B_2)^2)
\end{align}

@@

From now on we will apply equations \ref{eq:22} and \ref{eq:25}, to determine the compression efficiency of methods SM and VLB for random matrices generated as describe above.

With $w=0$, a single Beta distribution is used. In Figure \ref{fig:02},
we show some distributions of bit-lengths for some combinations of $\alpha_1$ and $\beta_1$. From the figure it can be seen that a large variety of unimodal distributions can be generated in the interval $[1,64]$.

\begin{figure}[ht]
  \centering
  \subfigure[$\alpha=1,\beta_1=1$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig02}
  }
  \subfigure[$\alpha=1,\beta=32$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig03}
  }
  \subfigure[$\alpha=32,\beta=1$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig04}
  }
  \subfigure[$\alpha=64,\beta=64$]{
  \includegraphics[scale=0.25,angle=-90]{fig05}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from a Beta distribution. Below each histogram is possible to verify the parameters used.}
  \label{fig:02}
\end{figure}

As we are sampling from a large set distributions of bit-length, represented by the mixture of betas presented above, in order to make our results more general, we will base our analysis on the expected bit-length of a sample, since the efficiency of both methods depends on it. So, from equations \ref{eq:22} and \ref{eq:25}, the expected efficiencies become:

\begin{equation}\label{eq:56}
 E(\eta_1) = 1 - \frac{k}{64}
\end{equation}

\begin{equation}\label{eq:57}
 E(\eta_2) = 1 - \frac{E(b)}{64} - \frac{k}{64}
\end{equation}

where k in \ref{eq:57}, is set to 7 (the bit-length required to represent the largest possible bit-length: 64). In \ref{eq:56}, $k$ is the bit-length of the greatest element, or in the worst case, 64.

We will use the difference $D=E(\eta_1)-E(\eta_2)$ to compare the efficiency of the two methods. Thus a positive $D$ will favor method $1$ while a negative $D$ favors method $2$.

@@

In Figure \ref{fig:03}, we can see the distribution of efficiencies and their difference for a sample generated from a single Beta distribution of bit-lengths. Note that both methods can achieve efficiencies greater than 80\% for matrices with very small numbers. Also note that the VLB method is more efficient in the majority of cases.

\begin{figure}[h]
  \centering
%   \subfigure[$\alpha=1,\beta_1=1$]{
%   \includegraphics[scale=0.28,angle=-90,clip]{fig02}
%   }
%   \subfigure[$\alpha=1,\beta=32$]{
%   \includegraphics[scale=0.28,angle=-90,clip]{fig03}
%   }
%   \subfigure[$\alpha=32,\beta=1$]{
%   \includegraphics[scale=0.28,angle=-90,clip]{fig04}
%   }
%   \subfigure[$\alpha=64,\beta=64$]{
%   \includegraphics[scale=0.28,angle=-90]{fig05}
%   }
  \includegraphics[scale=0.7]{simples.eps}
  \caption{Comparing compression efficiency of methods 1 and 2. Color scale in (a), (b) and (c) represent average bit-length. In (a) we can see the difference $D = \eta_1 - \eta_2$. It can be seen that for most combination of $\alpha$ and $\beta$, $D<0$, meaning the second method is more efficent to compress a sample of numbers with bit-lengths coming from a $Beta(\alpha,\beta)$ distribution. However, there is a small region in parameter space, which is shown in white on (d), where the first method is more efficient. This region corresponds to the dots in red in (a), where the average bit-length is higher. In panels (b) and (c), we can see the efficiencies of methods 1 and 2, respectively.}
  \label{fig:03}
\end{figure}

Now let $w = 0.5$, i. e., matrices will have elements with bit-lengths comming from a mixture of beta distributions, $B_1\sim Beta(\alpha_1,\beta_1)$ and $B_2\sim Beta(\alpha_2,\beta_2)$. The expected value for this mixture is shown in equation \ref{eq:58}.

\begin{equation}\label{eq:58}
 E(B) = 0.5 E(B_1) + 0.5 E(B_2)
\end{equation}

For bit-lengths coming from a mixture ($w>0$), let the expected efficiencies for the SM and VLB methods be as given by equations \ref{eq:59} and \ref{eq:60}. So now, instead of having the efficiency be a function of greatest bit-length in the sample (denoted as  $k$ in \ref{eq:56} and \ref{eq:57}), it will be a function of $max\{E(B_1),E(B_2)\}$.

\begin{equation}\label{eq:59}
 E(\eta_1) = 1 - \frac{max\{E(B_1),E(B_2)\}}{64}
\end{equation}

>>>>

Most if not all existing lossless compression methods were developed under the following usage paradigm: \textit{produce} $\rightarrow$ \textit{compress} $\rightarrow$ \textit{store} $\rightarrow$ \textit{uncompress} $\rightarrow$ \textit{use}. The focus of the present work is to allow a slightly different usage:  \textit{produce} $\rightarrow$ \textit{compress} $\rightarrow$ \textit{perform mathematical manipulations to use} and \textit{decompress} only for human reading.

@@

This paper will focus solely on numerical data which for the purpose of the applications is organized as   matrices. This is a most common data structure found in computational data analysis environments. The matrices compressed according to the methodologies proposed here should be able to undergo the same mathematical operations as the original uncompressed matrices. e.g. linear algebra manipulations. This way, the cost of compression is reduced to a single event of compression and no need of decompression except when displaying the results for human reading. The idea of operating with compressed arrays is relatively new\cite{yemliha2007compiler}, and it has yet to find mainstream applications to the field of numerical computations. One application which employs a form of compression is the sparse matrix linear algebra algorithms\cite{dodson1991sparse}, in this case there is no alteration in the standard encoding of the data, but only the non-zero elements of the matrices are stored and operated upon.

@@

To maintain mathematical equivalence with the original data for any arithmetic operations, we need to  maintain the structure of the matrix, i.e., the ability to acess any element given its row $i$ and column $j$ and also the numeric nature of its elements. In order to achieve compression we decided to exploit inefficiencies in the conventional way matrices are allocated in memory. The examples in this paper will be restricted to matrices with positive integer elements.

The compression method is as follows. Let $M_{r \times c}$ be a matrix, in which $r$ is the number of rows and $c$ the number of columns. Each element of this matrix, called $m_{ij}$, is a positive integer. In digital computers, all information is stored as binary code (base 2 numbers). However the conventional way to store arrays of integers is on a memory block sequence of fixed size (power of 2 bit numbers), one for each element. The maximum size of a block is equal to the word size of the processor, which for most current CPUs is 64 bits. Some special number such as complex number may be encoded as two blocks instead of one. The size of the chunk of memory allocated to each number will determine their maximum  size (for integers) or their precision (for floating-point numbes). So for matrix $M$, the total memory allocated, assuming chunks of 64 bits, is given by $\mathcal{B} = r \times c \times 64$.

@@

The SM method consists in determining the value of the greatest element of matrix $M$, which coincides with its supremum, $max M == \sup M$  and determine the minimum bit number, $b(\sup M$)(Equation \ref{eq:01}), required to store it. We will use capital roman letters to denote uncompressed matrices and the corresponding lower case letter for the compressed version.

\begin{equation} \label{eq:01}
 b(\sup M) \approx \begin{cases}
	1, &  \text{ if } \sup M \in \{0,1\} \\ 
	\lfloor \log_2(\sup M)  \rfloor + 1,  & \text{ if } \sup M > 1  
	\end{cases}
\end{equation}

The allocation of memory still happens in the usual way, i.e., in fixed size 64-bit chunks, only that now, in the space required for a single 64 bit integer, we can store for example, an entire $8\times 8$ matrix of $0_{10}$ and $1_{10}$.

Let's look at a concrete example: suppose that the greatest value to be stored in a matrix $M$ is $\max M=1023$. Therefore, the bit number required to represent it is $10$ $(1111111111)$. Let the first 8 elements of $M$ be:

\begin{equation} \label{eq:02}
  M = \begin{bmatrix}
  900 & 1023 & 721 & 256 & 1 & 10 & 700 & 20 & \hdots\\ 
  \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \ddots 
  \end{bmatrix}
\end{equation}

These elements of $M$, in binary, are shown in Table \ref{tab:01}. It is evident that the bit number required to represent any other element must be lower or equal to $10$. From now on the minimum number of bits required to represent a base 10 integer will be refered to as its bit-length.

@@

To store matrix $M$ it first has to be converted to base $2$ ($M_2$). Then it will be unraveled by column (column major, e.g. in Fortran) or by row (row major, e.g. in C) and its elements will be written as  fixed size adjacent chunks of memory. The size of each chunk is determined by the type associated with the matrix (typically 64 bits, but always a power of 2).

According to the SM method, having determined that each element will require at most 10 bits, we can divide the memory block corresponding to a single 64 bit integer into six 10-bit chunks which can each hold a single element of $M$. These 64-bit blocks will be called a bitstring. The remaining 4 bits will be used later. The number of bitstrings needed will be $\lfloor \frac{\dim(M) *b(\sup M)} {64} \rfloor +1$.

The final layout of the first 6 elements of $m$ in the first bitstring can be seen in \ref{eq:03}.

\begin{equation}\label{eq:03}
 bitstring_1 = 0000\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\underbrace{0100000000}_{256}\underbrace{1011010001}_{721}\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
\end{equation}

Here is a step-by-step description of the application of the SM method to matrix $M$:

\begin{enumerate}
 \item Element $M_{1,1}=900 = m_{1,1} = 1110000100$ is stored in the first 10-bit chunk of the element strip $bitstring[1]$ , which corresponds to bits 0 to 9.
\begin{equation*}\label{eq:04}
bitstring_1 = 000000000000000000000000000000000000000000000000000000\underbrace{1110000100}_{900} 
\end{equation*}
 \item Element $M_{1,2}=1023$ is allocated in the second chunk, from bit 10 to bit 19.
\begin{equation*}\label{eq:05}
 bitstring_1 = 00000000000000000000000000000000000000000000\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
\end{equation*}
 \item Repeat for elements $M_{1,i}$ with $i=1,\ldots,6$ which are stored on the remaining chunks.
 \begin{equation*}\label{eq:06}
bitstring_1 = 0000\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\underbrace{0100000000}_{256}\underbrace{1011010001}_{721}\underbrace{1111111111}_{1023}\underbrace{1110000100}_{900} 
 \end{equation*}
 \item Element 7 does not fit on the remaining 4 bits of the first bitstring. So it will straddle two bitstrings, i.e., it is divided in two segments $a$ and $b$, $a$ is written on the first bitstring and $b$ on the second.
 \begin{equation*}\label{eq:07}
  \underbrace{\underbrace{0000010100}_{20}\overbrace{101011}^{b}}_{bitstring_2}|\underbrace{\overbrace{1100}^{a}\underbrace{0000001010}_{10}\underbrace{0000000001}_{1}\hdots}_{bitstring_1}
\end{equation*}
\end{enumerate}

@@

\begin{equation*}\label{eq:8}
  \underbrace{0000010100}_{20}\underbrace{\overbrace{101011}^a|\overbrace{1100}^{b}}_{700}\underbrace{0000001010}_{10}\ldots
\end{equation*}

@@

Lets use again the matrix described in Equation \ref{eq:02}, where  the largest element is number 1023. Now instead of assigning one chunk of the bitstring to each element of $m$, we will assign two chunks: the first will store the number of bits required to store the element and the second  will store the actual element. The first chunk will have a fixed size, in this case, 4 bits. These 4 bits are the required space to store the bit-length of $\sup M $, in this case, 10.

@@

\begin{enumerate}
 \item The first element $M_{1,1}=900$ requires 10 bits to store, so we write $10$ in the first chunk and $900$ in the second.
 \begin{equation*} \label{eq:09}
 bitstring_1 = 00000000000000000000000000000000000000000000000000\underbrace{\underbrace{1110000100}_{\text{element} = 900}\underbrace{1010}_{\text{bit-length=10}}}_{M_{1,1}}
\end{equation*}
 \item Do the same for the next element, $M_{1,2}=1023$.
  \begin{equation*} \label{eq:10}
 bitstring_1 = \ldots00000000000000000000000000\underbrace{\underbrace{1111111111}_{\text{element} = 1023}\underbrace{1010}_{\text{bit-length=10}}}_{M_{1,2}}\underbrace{\underbrace{1110000100}_{\text{element} = 900}\underbrace{1010}_{\text{bit-length=10}}}_{M_{1,1}}
\end{equation*}
 \item Element $M_{1,3}=721$ is also added taking the bitstring to the state.
 \begin{equation*} \label{eq:11}
 bitstring_1 = 
 0000000000000000000000\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10}
\end{equation*}
\end{enumerate}

@@

\begin{enumerate}
 \item[4.] Element $M_{1,4} =256$ is added.  
 \item [5.] Elements $M_{1,5} =1$ and $M_{1,6} =10$ are added requiring a total of 13 bits instead of 20 with the SM method. Now, it is used 2 bit strips.
 \begin{align*} \label{eq:12}
 bitstring_1 &=
 \underbrace{0100}_{4}\underbrace{1}_{1}\underbrace{0001}_{1}\underbrace{100000000}_{256}\underbrace{1001}_{9}\underbrace{1011010001}_{721}\underbrace{1010}_{10}\underbrace{1111111111}_{1023}\underbrace{1010}_{10}\underbrace{1110000100}_{900}\underbrace{1010}_{10} \\
 bitstring_2 &= 000000000000000000000000000000000000000000000000000000000000\underbrace{1010}_{10}
\end{align*}
 \item [6.] The remaining two elements are added $M_{1,7} =700$ and $M_{1,8} =20$ in the second bit strip.
 \begin{equation*} \label{eq:13}
 bitstring_2 = 00000000000000000000000000000000000\underbrace{10100}_{20}\underbrace{0101}_{5}\underbrace{1010111100}_{700}\underbrace{1010}_{10}\underbrace{1010}_{10}
\end{equation*}
\end{enumerate}

@@

 \begin{equation}\label{eq:14}
  \eta=\frac{bits\, alocated-bits\, used}{bits\,alocated}
 \end{equation}

 Where \textit{bits alocated} above mean total bits required for standard storage of the matrix, without compression, while bits used mean total bits requires to store the matrix after compression. From now on the efficiencies are denoted by $\eta_1$ for the SM method and by $\eta_2$ for the VLB method. 
 \subsubsection*{SM Method}
 Let $M_{r \times c}$ be the matrix we wish to compress. In comparison with a conventional allocation (64-bit integers) we can apply Equation \ref{eq:14} to caluculate the efficiency of the SM method:

\begin{align}\label{eq:15}
 \eta_1 &= \frac{64 \times rc - b(max M) \times rc}{64 \times rc}\nonumber\\
 &= \frac{64  - b(max M) }{64}
\end{align}

As we see in \ref{eq:15}, $\eta_1$ does not depend on size of the matrix, only on the bit-length of $\max M$. If $b(\max M)=64$, $\eta_1$ is 0, i. e., no compression is possible. On the other extreme, if the matrix is composed exclusively of 0s and 1s, maximal compression is achievable, $\eta_1=1$.

@@

\begin{equation}\label{eq:16}
 \eta_2 = \frac{64 \times rc - \sum_{i=1}^{g} ( b_i + k ) \times f_i }{64 \times rc} 
\end{equation}

We can further simplify Equation \ref{eq:16} to get at shorter expression for the compression ratio.

\begin{align} \label{eq:17}
 \eta_2 &= \frac{64 \times rc - \sum_{i=1}^{g} ( b_i \times f_i + k \times f_i )}{64 \times rc} \nonumber \\
  &= \frac{64 \times rc - \sum_{i=1}^{g}  b_i \times f_i  -\sum_{i=1}^{g}  k \times f_i }{64 \times rc}\nonumber \\
  &= \frac{64 \times rc - \sum_{i=1}^{g}  b_i \times f_i  - k \times\sum_{i=1}^{g}  f_i }{64 \times rc}\nonumber \\
 &\textrm{Please note that }\sum_{i=1}^{g} f_i = rc, \nonumber \\
 \eta_2 &= \frac{64 \times rc - \sum_{i=1}^{g}  b_i \times f_i  - k \times rc }{64 \times rc}\nonumber \\
  &= 1 - \frac{\sum_{i=1}^{g}  b_i \times f_i }{64 \times rc} - \frac{k}{64}
\end{align}

@@

\paragraph{Random Matrix Generation.}
In both methods, compression efficiency depends on the distribution of the bit-lengths $b(m_{i,j})$. Thus, in this section, a method to generate a variety of random bit-length distributions is proposed.

For simplicity we will model the distribution of  as a mixture $X$ of two Beta distributions, $B_1 \sim Beta(\alpha_1,\beta_1)$ and $B_2 \sim Beta(\alpha_2,\beta_2)$, whose probability function is shown in Equation \ref{eq:18}. Since the Beta distribution is defined only in the interval $[0,1] \subset \mathbb{R}$ , we applied a simple transformation ($\lfloor 64 \times x \rfloor + 1$) to the mixture in order to map it to the interval of $[1,64] \subset \mathbb{Z}$.

\begin{equation}\label{eq:18}
 F(x) =  w \, Beta(\alpha_1,\beta_1) + (1-w)\,Beta(\alpha_2,\beta_2)
\end{equation}

The intention of using this mixture was to find a simple way to represent a large variety of bit length distributions. The first two central moments of this mixture are given in \ref{eq:19} and will be used later to summarize our numerical results.

\begin{align}\label{eq:19}
 E(X) &= w E(B_1) + (1-w) E(B_2) \nonumber \\
 &= w \frac{\alpha_1}{\alpha_1+\beta_1} + (1-w)\frac{\alpha_2}{\alpha_2+\beta_2}\nonumber \\
 Var(X) &= w Var(B_1) + (1-w) Var(B_2) + w (1-w) (E(B_1)^2 - E(B_2)^2)
\end{align}

@@

From now on we will apply Equations \ref{eq:15} and \ref{eq:17}, to determine the compression efficiency of SM and VLB methods for random matrices generated as describe above.

With $w=0$, a single Beta distribution is used. In Figure \ref{fig:01020304}, we show some distributions of bit-lengths for some combinations of $\alpha_1$ and $\beta_1$. From the figure it can be seen that a large variety of unimodal distributions can be generated in the interval $[1,64]$.

\begin{figure}[ht]
  \centering
  \subfigure[$\alpha=1,\beta_1=1$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig01}
  }
  \subfigure[$\alpha=1,\beta=32$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig02}
  }
  \subfigure[$\alpha=32,\beta=1$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig03}
  }
  \subfigure[$\alpha=64,\beta=64$]{
  \includegraphics[scale=0.25,angle=-90]{fig04}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from a Beta distribution. Below each histogram is possible to verify the parameters used.}
  \label{fig:01020304}
\end{figure}

As we are sampling from a large set distributions of bit-length, represented by the mixture of betas presented above, in order to make our results more general, we will base our analysis on the expected bit-length of a sample, since the efficiency of both methods depends on it. So, from Equations \ref{eq:15} and \ref{eq:17}, the expected efficiencies become:

\begin{equation}\label{eq:20}
 E(\eta_1) = 1 - \frac{k}{64}
\end{equation}

\begin{equation}\label{eq:21}
 E(\eta_2) = 1 - \frac{E(b)}{64} - \frac{k}{64}
\end{equation}

\noindent where k in \ref{eq:21}, is set to 7 (the bit-length required to represent the largest possible bit-length: 64). In \ref{eq:20}, $k$ is the bit-length of the greatest element, or in the worst case, 64.

We will use the difference $D=E(\eta_1)-E(\eta_2)$ to compare the efficiency of the two methods. Thus a positive $D$ will favor SM method while a negative $D$ favors VLB method.

@@

In Figure \ref{fig:06070809}, we can see the distribution of efficiencies and their difference for a sample generated from a single Beta distribution of bit-lengths. Note that both methods can achieve efficiencies greater than 80\% for matrices with very small numbers. Also note that the VLB method is more efficient in the majority of cases.

\begin{figure}[h]
  \centering
   \subfigure[SM-VLB  Difference ($D = \eta_1 - \eta_2$)]{ 
   \includegraphics[scale=0.18,clip]{fig07}
   }
   \subfigure[SM Efficiency  ($\eta_1$)]{
   \includegraphics[scale=0.18,clip]{fig06}
   }
   \subfigure[VBL Efficiency ($\eta_2$)]{
   \includegraphics[scale=0.18,clip]{fig05}
   }
   \subfigure[SM$\geq$VLB Efficiency ($\eta_1 \geq \eta_2$)]{
   \includegraphics[scale=0.18,clip]{fig08}
   }
  \caption{Comparing compression efficiency of methods 1 and 2. Color scale in (a), (b) and (c) represent average bit-length. In (a) we can see the difference $D = \eta_1 - \eta_2$. It can be seen that for most combination of $\alpha$ and $\beta$, $D<0$, meaning the second method is more efficent to compress a sample of numbers with bit-lengths coming from a $Beta(\alpha,\beta)$ distribution. However, there is a small region in parameter space, which is shown in white on (d), where the first method is more efficient. This region corresponds to the dots in red in (a), where the average bit-length is higher. In panels (b) and (c), we can see the efficiencies of SM and VLB methods, respectively.}
  \label{fig:06070809}
\end{figure}

Now let $w = 0.5$, i.e., matrices will have elements with bit-lengths comming from a mixture of beta distributions, $B_1\sim Beta(\alpha_1,\beta_1)$ and $B_2\sim Beta(\alpha_2,\beta_2)$. The expected value for this mixture is shown in Equation \ref{eq:22}.

\begin{equation}\label{eq:22}
 E(B) = 0.5 E(B_1) + 0.5 E(B_2)
\end{equation}

For bit-lengths coming from a mixture ($w>0$), let the expected efficiencies for the SM and VLB methods be as given by Equations \ref{eq:23} and \ref{eq:24}. So now, instead of having the efficiency be a function of greatest bit-length in the sample (denoted as  $k$ in \ref{eq:15} and \ref{eq:17}), it will be a function of $max\{E(B_1),E(B_2)\}$.

\begin{equation}\label{eq:23}
 E(\eta_1) = 1 - \frac{max\{E(B_1),E(B_2)\}}{64}
\end{equation}

\begin{equation}\label{eq:24}
 E(\eta_2) = 1 - 0.5\frac{E(B_1)}{64} - 0.5\frac{E(B_2)}{64} - \frac{max\{E(B_1),E(B_2)\}}{64}
\end{equation}

As before, we generate 3 matrices of dimension $10,000$ for each parameterization, calculate the average efficiencies (Equations \ref{eq:23} and \ref{eq:24}) and their diference $D$.
